Planetary perturbers: Flaring star-planet interactions in Kepler and TESS

### 2023-10-10 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract with filler text
\begin{abstract}
    In many star-planet systems discovered so far, the innermost planet orbits within only a few stellar radii. In these systems, planets \textcolor{red}{could} become in-situ probes of the extended stellar magnetic field. Because they disturb the field as they move, they are expected to trigger flares on the stellar surface. These flares are not expected to appear different from intrinsic stellar flares individually. However, as we expect planet-induced flares to be associated with the planet's orbital period $P_{\rm orb}$, we can identify their presence from a clustering of flares in phase with the planet's orbital phase. We used the excellent phase coverage obtained from Kepler and the Transiting Exoplanet Survey Satellite to find flaring star-planet systems, compile a catalog of all their flares, and measure how much they cluster in orbital phase.

    In the 1811 searched systems, we found 25 single stars with three or more flares each. We quantified the significance of the clustering in each system, and compared it against the theoretically expected power of magnetic interaction that leads to planet-induced flaring. Most systems do not show any clustering, in accordance with low expected power. Those we expect to show planet-induced flaring fall on two branches. An inactive one, without any signs of clustering, and a tentative active one, where the clustering becomes more pronounced as the expected power of interaction increases. The flares in HIP 67522, a young Hot Jupiter system, are prominently clustered ($p<0.006$), identifying it as the most promising candidate for magnetic star-planet interaction in our sample.

@@

The absolute power of flaring SPI is difficult to constrain. We do not know how energetic planet-induced flares can be. The estimates for dissipated energy~\citep{lanza2018closeby} are in the ballpark of the regime of superflares observed by Kepler and TESS. However, on the one hand, the planet might be triggering flares prematurely, which would be of lower energy~\citep{loyd2023flares}, and end up below our detection threshold. On the other hand, the energy need not be dissipated continuously, particularly in the stretch-and-break mechanism proposed by~\citep{lanza2012starplanet}, so an individual flare's energy might be higher. We also do not know whether flares triggered by interactions with a companion would have different properties than intrinsic flares. We will assume that, since they would be caused by magnetic reconnection triggered by the planet in the stellar corona, they should still have the same physics as intrinsic flares.

Since planet-induced flares might not look any different from intrinsic flares, and we do not know how their energies will be distributed, we require a method to distinguish them without referring to their individual properties. Instead, we can look at the statistical distribution of flares in time:

@@

\subsection{Searching for local changes in activity}
\label{sec:intro:local}
In contrast to global SPI, local SPI effects are easier to tell apart from intrinsic stellar behavior because they are tied to the planet's orbital period. 
Phase-correlated activity, flaring and otherwise, has been searched for in individual systems. After eight years of monitoring of HD 189733, an active Hot Jupiter host, higher X-ray energy flares seem only marginally clustered in orbital phase than lower energy ones~\citep{pillitteri2022xray}. \citet{maggio2015coordinated} observed an eruption of X-ray emission when the Jupiter-sized companion was in periastron on the highly eccentric orbit around its host HD 17156. However, a similar study using chromospheric indicators in the eccentric system HD 80606 found no variation between periastron and apoastron~\citep{figueira2016activity}. In the above cases, non-detection can at least partly be attributed to poor orbital phase coverage, which also limits the significance of any positive detection. Realizing this, \citet{fischer2019timevariable} used K2 data with excellent phase coverage of the orbits of the inner \mbox{TRAPPIST-1} planets to look for clustered flaring in phase, but found none. \citet{ilin2022searching} and \citet{klein2022one} found hints of excess flaring, and chromospheric Helium emission in the young AU Mic system, respectively. This mixed picture of individual object studies calls for a more comprehensive and systematic approach.

@@

\begin{longrotatetable}
\script{paper_main_table.py}
\movetableright=-20mm
    % \caption{Flaring single star-planet system parameters. Obs. time is the total observing time with Kepler and TESS.}
% \footnotesize
    \input{output/table_lit_vals.tex}
        \label{tab:maintable_lit}
    \tablerefs{\input{output/lit_table_bibstring.tex}}
\end{longrotatetable}

@@

>>>>

% Abstract with filler text
\begin{abstract}
    In many star-planet systems discovered so far, the innermost planet orbits within only a few stellar radii. In these systems, planets \textcolor{red}{could} become in-situ probes of the extended stellar magnetic field. Because they disturb the field as they move, they are expected to trigger flares on the stellar surface. \textcolor{red}{If and how their energy and morphology differs from intrinsic stellar flares is poorly constrained.} However, as we expect planet-induced flares to be associated with the planet's orbital period $P_{\rm orb}$, we can identify their presence from a clustering of flares in phase with the planet's orbital phase. We used the excellent phase coverage from Kepler and the Transiting Exoplanet Survey Satellite to find flaring star-planet systems, compile a catalog of all their flares, and measure how much they cluster in orbital phase.

    In the 1811 searched systems, we found 25 single stars with three or more flares each. We quantified the significance of the clustering in each system, and compared it against the theoretically expected power of magnetic interaction that leads to planet-induced flaring. Most systems do not show any clustering, consistent with low expected power. Those we expect to show planet-induced flaring fall on two branches. An inactive one, without any signs of clustering, and a tentative active one, where the clustering becomes more pronounced as the expected power of interaction increases. The flares in HIP 67522 are prominently clustered ($p<0.006$). This young Hot Jupiter system is the most promising candidate for magnetic star-planet interaction in our sample.

@@

The absolute power of flaring SPI is difficult to constrain. We do not know how energetic planet-induced flares can be. The estimates for dissipated energy~\citep{lanza2018closeby} are in the ballpark of the regime of superflares\textcolor{red}{, with energies released at optical wavelengths $\geq10^{33}\,$erg~\citep{schaefer2000superflares},} observed by Kepler and TESS. However, on the one hand, the planet might be triggering flares prematurely, which would be of lower energy~\citep{loyd2023flares}, and end up below our detection threshold. On the other hand, the energy need not be dissipated continuously, particularly in the stretch-and-break mechanism proposed by~\citep{lanza2012starplanet}, so an individual flare's energy might be higher. We also do not know whether flares triggered by interactions with a companion would \textcolor{red}{unfold faster or slower than} intrinsic flares. We will assume that, since they would be caused by magnetic reconnection triggered by the planet in the stellar corona, they should still have the same physics as intrinsic flares\textcolor{red}{. They should then approximalety follow the fast-rise-exponential-decay morphology. Individual flares can deviate from this template, yet still be classified as flares because flares remain the best explanation for a sudden rise and drop in stellar brightness in most main sequence dwarf stars. Many of these deviations are unexplained today~\citep{howard2022no}, so that identifying SPI flares based on morphology is not possible without additional constraints. However, if we can statistically identify enough SPI flares, as we attempt in this work, we can infer the differences between intrinsic and SPI events from the two ensembles.}

\textcolor{red}{In other words, s}ince planet-induced flares might not look any different from intrinsic flares, and we do not know how their energies will be distributed, we require a method to distinguish them without referring to their individual properties. Instead, we can look at the statistical distribution of flares in time:

@@

\subsection{Searching for local changes in activity}
\label{sec:intro:local}
In contrast to global SPI, local SPI effects are easier to tell apart from intrinsic stellar behavior because they are tied to the planet's orbital period. \textcolor{red}{Searches for star-planet interactions in polarized radio emission, i.e. at the wavelengths at which Jupiter-Io interaction is observed (scaled up to the high field strengths of stars) have produced promising, but tentative results. In some works, shorter radio wavelenths can still be confused with stellar activity~\citep{pineda2023coherent}. In others, the planet has not been detected (yet)~\citep{vedantham2020coherent}, the data do not cover multiple orbital periods to see the signal repeat~\citep{perez-torres2021monitoring}, or both \citep{callingham2021population}.} Phase-correlated activity, flaring and otherwise, has been searched for in individual systems. After eight years of monitoring of HD 189733, an active Hot Jupiter host, higher X-ray energy flares seem only marginally clustered in orbital phase than lower energy ones~\citep{pillitteri2022xray}. \citet{maggio2015coordinated} observed an eruption of X-ray emission when the Jupiter-sized companion was in periastron on the highly eccentric orbit around its host HD 17156. However, a similar study using chromospheric indicators in the eccentric system HD 80606 found no variation between periastron and apoastron~\citep{figueira2016activity}. In the above cases, non-detection can at least partly be attributed to poor orbital phase coverage, which also limits the significance of any positive detection. Realizing this, \citet{fischer2019timevariable} used K2 data with excellent phase coverage of the orbits of the inner \mbox{TRAPPIST-1} planets to look for clustered flaring in phase, but found none. \citet{ilin2022searching} and \citet{klein2022one} found hints of excess flaring, and chromospheric Helium emission in the young AU Mic system, respectively. This mixed picture of individual object studies calls for a more comprehensive and systematic approach.

@@

\begin{table*}
\begin{rotatetable*}
\script{paper_main_table.py}
% \rotate
 \caption{Flaring single star-planet system parameters. Obs. time is the total observing time with Kepler and TESS.}
\movetableright=-40mm

\footnotesize

    \input{output/table_lit_vals.tex}
        \label{tab:maintable_lit}
    \tablerefs{\input{output/lit_table_bibstring.tex}}
\end{rotatetable*}
\end{table*}

@@

% \textcolor{red}{Mid-to-late M dwarfs follow this bimodal distribution }
